### 2014-11-05 contract-language.tex
edits

% Contract Language Syntax
% ------------------------
\input{contract-syntax.tex}

In this section, we formalize \name's contract language, and describe our
contract classification scheme, which analyzes a contract and maps it to the
weakest store consistency level sufficient to satisfy its consistency
requirements.

% classifies contracts on the basis of the weakest
% store-level consistency guarantee

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {\eff} \in \mathtt{Effect} \qquad
  {\cv} \in \mathtt{Contract} \qquad
  \set{\eff} \in \mathtt{Effect\; Set}
}\\
\EffSoup & \in & \mathtt{EffSoup}	  & \coloneqq & \set{\eff} \\
\visZ, \soZ, &	\in & \mathtt{Relations} & \coloneqq &
  \set{\eff}\times\set{\eff} \\
\sameobjZ		&     &  & \\
{\E} 		& \in & \mathtt{ExecState}  & \coloneqq & \Exec \\
\end{array}
\end{smathpar}

@@

An important aspect of our contract language is its ability to capture
store-level consistency guarantees, along with application-level consistency
requirements. Similar to~\cite{Burckhardt2014}, we can rigorously define a wide
variety of store semantics including those that combine any subset of session
and causality guarantees, and multiple consistency levels.  However, for our
purposes, we identify three particular consistency levels -- eventual, causal,
and strong, commonly offered by many distributed stores with tunable
consistency, with increasing overhead in terms of latency and availability. For
each of the these three consistency levels, we capture the semantics as a
formula in our contract language, and informally describe the kind of
application-level consistency requirements that can be met under the
consistency level:

% Indeed, the techniques presented here can be
% extended to other consistency stratifications. We assign each
% application-level contract into one of these following classes:

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
	be satisfied as long as the client can reach at least one replica. For
	example, \cf{deposit} is an eventually consistent operation; its semantics
	does not require its action to manifest on all replicas before other
	operations in its session are allowed to proceed. While eventually consistent
	data stores typically offer \emph{basic} eventual consistency with all
	possible anomalies, we assume that our store provides stronger semantics that
	remain highly-available~\cite{BailisHAT,COPS}; the store always exposes a
	causal cut of the updates. This semantics can be formally captured in terms
	of the following contract definition:

  \vspace{-1em}
  \begin{smathpar}
  \ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
  \end{smathpar}
  \noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Causally consistent operations are required
	to see a causally consistent snapshot of the object state, including the
	actions performed on the same session.  The latter requirement implies that
	if two operations $o_1$ and $o_2$ from the same session are applied to two
	different replicas $r_1$ and $r_2$, the second operation cannot be discharged
	until the effect of $o_1$ is merged with $o_2$ in both $r_1$ and $r_2$. The
	\cf{getBalance} operation requires causal consistency, as it requires the
	operations from the same session to be visible, which cannot be guaranteed
	under just eventual consistency. The corresponding store semantics is
	captured by the contract $\ccc$ defined below:

@@

  %%; two operations in the same session but on different objects are
  %% considered causally unrelated under this definition. Stores typically
  %%avoid tracking causality across objects to mitigate overheads when
  %% causality tracking is unnecessary.

@@

Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves totally
ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$.  This
concurs with the intuition that any contract satisfiable under $\ecc$ or $\ccc$
is satisfiable under $\scc$, and any contract that is satisfiable under $\ecc$
is satisfiable under $\ccc$. However, we are interested in the \emph{weakest}
guarantee (among $\ecc$, $\ccc$, and $\scc$) required to satisfy the contract.
We define the corresponding consistency level as the \emph{consistency class of
the contract}. The classification scheme, presented formally in
Figure~\ref{sem:classify}, defines rules to judge the consistency class of a
contact. For example, the scheme classifies the \cf{getBalance} contract
($\cv_{gb}$) from \S~\ref{sec:motivation} as a {\sf CausallyConsistent}
contract, because the sequent $\Delta \vdash \ccc \Rightarrow \cv_{gb}$ is
valid in first-order logic (therefore, $\cv_{gb} \le \ccc$), whereas the
sequent $\Delta \vdash \ecc \Rightarrow \cv_{gb}$ is invalid (therefore,
$\cv_{gb} \not\le \ecc$). Due to the confinement of our contract language to a
decidable subset of the logic, validity of such sequents can be decided
mechanically allowing us to automate the classification scheme in \name.

Along with three straightforward rules that classify contracts into consistency
classes, the classification scheme also presents a rule that judges
well-formedness of a contract. A contract is well-formed if and only if it is
satisfiable under $\scc$ - the strongest possible consistency guarantee any
store can provide. Otherwise, it is considered ill-formed, and rejected
statically.

@@

\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {op} \in \mathtt{Operation} \qquad
  \tau \in \mathtt{Consistency\; Class}
}\\
\cv(\tau) & \in & \mathtt{Store\; Contract\; of \; \tau} & \coloneqq & \scc,
  \ccc, \ecc\\
{\sigma} & \in & \mathtt{Session} & \coloneqq & \cdot \ALT (op,\tau); \sigma \\
\Sigma 	& \in & \mathtt{Session\;Soup}   & \coloneqq &
  \langle s,{\sigma} \rangle \pll \Sigma \ALT \emptyset \\
				&			&	\mathtt{Store}		  & \coloneqq & \E;\Sigma \\
\end{array}
\end{smathpar}

We model the system as a tuple $\E;\Sigma$, where the execution $\E$
captures the data store's current state of the execution, and
$\Sigma$, the session soup, is the set of concurrent client sessions
interacting with the store. A session is modeled simply as a sequence
of replicated data type operations ($\mathit{op}$), tagged with the consistency
class ($\tau$) of their contracts (as determined by the contract
classification scheme). We assume a rewrite relation of form:

>>>>

\input{contract-syntax.tex}

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {\eff} \in \mathtt{Effect} \qquad
  {\cv} \in \mathtt{Contract} \qquad
  \set{\eff} \in \mathtt{Effect\; Set}
}\\
\EffSoup & \in & \mathtt{EffSoup}	  & \coloneqq & \set{\eff} \\
\visZ, \soZ, \sameobjZ &	\in & \mathtt{Relations} & \coloneqq & \EffSoup \times \EffSoup \\
%\sameobjZ		&     &  & \\
{\E} 		& \in & \mathtt{ExecState}  & \coloneqq & \Exec \\
\end{array}
\end{smathpar}

@@

An important aspect of our contract language is its ability to capture
store-level consistency guarantees, along with application-level consistency
requirements. Similar to~\cite{Burckhardt2014}, we can rigorously define a wide
variety of store semantics including those that combine any subset of session
and causality guarantees, and multiple consistency levels.  However, for our
purposes, we identify three particular consistency levels -- eventual, causal,
and strong, commonly offered by many distributed stores with tunable
consistency, with increasing overhead in terms of latency and availability.

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
	be satisfied as long as the client can reach at least one replica. In the
	bank account example, \cf{deposit} is an eventually consistent operation.
	While eventually consistent data stores typically offer \emph{basic} eventual
	consistency with all possible anomalies, we assume that our store provides
	stronger semantics that remain highly-available~\cite{BailisHAT,COPS}; the
	store always exposes a \emph{causal cut} of the updates. This semantics can
	be formally captured in terms of the following contract definition:

  \vspace{-1em}
  \begin{smathpar}
  \ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
  \end{smathpar}

\item \textbf{Causal consistency}: Causally consistent operations are required
	to see a causally consistent snapshot of the object state, including the
	actions performed on the same session.  The latter requirement implies that
	if two operations $o_1$ and $o_2$ from the same session are applied to two
	different replicas $r_1$ and $r_2$, the second operation cannot be discharged
	until the effect of $o_1$ is included in $r_2$. The \cf{getBalance} operation
	requires causal consistency, as it requires the operations from the same
	session to be visible, which cannot be guaranteed under eventual consistency.
	The corresponding store semantics is captured by the contract $\ccc$ defined
	below:

@@

@@

Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves totally
ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$.  This
concurs with the intuition that any contract satisfiable under $\ecc$ or $\ccc$
is satisfiable under $\scc$, and any contract that is satisfiable under $\ecc$
is satisfiable under $\ccc$. We are interested in the \emph{weakest} guarantee
(among $\ecc$, $\ccc$, and $\scc$) required to satisfy the contract. We define
the corresponding consistency level as the \emph{consistency class} of the
contract.

The classification scheme, presented formally in Figure~\ref{sem:classify},
defines rules to judge the consistency class of a contact. For example, the
scheme classifies the \cf{getBalance} contract ($\cv_{gb}$) from
\S~\ref{sec:motivation} as a {\sf\small CausallyConsistent} contract, because
the sequent $\Delta \vdash \ccc \Rightarrow \cv_{gb}$ is valid in first-order
logic (therefore, $\cv_{gb} \le \ccc$), whereas the sequent $\Delta \vdash \ecc
\Rightarrow \cv_{gb}$ is invalid (therefore, $\cv_{gb} \not\le \ecc$). Since we
confine of our contract language to a decidable subset of the logic, validity
of such sequents can be decided mechanically allowing us to automate the
classification scheme in \name.

Along with three straightforward rules that classify contracts into consistency
classes, the classification scheme also presents a rule that judges
well-formedness of a contract. A contract is well-formed if and only if it is
satisfiable under $\scc$ - the strongest possible consistency guarantee that
the store can provide. Otherwise, it is considered ill-formed, and rejected
statically.

@@

\vspace{-1em}
\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{c}{
  {op} \in \mathtt{Operation} \qquad
  \tau \in \mathtt{Consistency\; Class}
}\\
\cv(\tau) & \in & \mathtt{Store\; Contract\; of \; \tau} & \coloneqq & \scc,
  \ccc, \ecc\\
{\sigma} & \in & \mathtt{Session} & \coloneqq & \cdot \ALT (op,\tau); \sigma \\
\Sigma 	& \in & \mathtt{Session\;Soup}   & \coloneqq &
  \langle s,{\sigma} \rangle \pll \Sigma \ALT \emptyset \\
				&			&	\mathtt{Config}		  & \coloneqq & \E;\Sigma \\
\end{array}
\end{smathpar}

We model the system as a tuple $\E;\Sigma$, where the axiomatic execution $\E$
captures the data store's current state, and session soup $\Sigma$ is the
set of concurrent client sessions interacting with the store. A session is
modeled simply as a sequence of replicated data type operations
($\mathit{op}$), tagged with the consistency class ($\tau$) of their contracts
(as determined by the contract classification scheme). We assume a rewrite
relation of form:
